---
title: "Activity: Bird Species"
author: "Vydeepthi Dhulipalla"
date: "June 7,2023 "
output: html_notebook
---


# Set Up

```{r}
# Load `tidyverse` and `dcData` packages

library(tidyverse)
library(dcData)

# Load the `OrdwayBirds` data set into your RStudio environment from the `dcData` package

data(OrdwayBirds, package = "dcData")

# Show a few commands used to inspect the data set a few different ways

glimpse(OrdwayBirds)
View(OrdwayBirds)

```

# Activity

When you're finished with this activity, you will have a graph that shows what time of year various species appear at the Katherine Ordway Natural History Study Area in Inver Grove Heights, MN.

### Step 0

Before we begin, the book gives us some instructions to select a few key variables and clean up the date formatting.  Use the commands given in the book, but be sure to study the functions used so you know what is happening.  In this case, we select `SpeciesName`, `Month`, and `Day`. Then a `mutate()` command converts each variable to character, and then converts the resulting "character string" to a number.

```{r}
# Get the data table & clean up dates (see Data Computing eBook)
OrdwayBirds_cleaned <- OrdwayBirds %>%
  mutate(
    Month = as.numeric(Month),
    Day = as.numeric(Day)
  )

# Inspect resulting data
head(OrdwayBirds_cleaned)

```



### Step 1




```{r}
# unique species in original data (including mis-spellings)

unique_species_original <- unique(OrdwayBirds$SpeciesName)

# Print unique species in original data
unique_species_original



# unique species in the clean list
unique_species_cleaned <- unique(OrdwayBirds_cleaned$SpeciesName)

# Print unique species in the clean list
unique_species_cleaned


```


### Step 2
  



```{r}
# Inspect OrdwaySpeciesNames data
View(OrdwaySpeciesNames)

# Inner join
joined_data <- inner_join(OrdwayBirds_cleaned, OrdwaySpeciesNames, by = "SpeciesName")

# Inspect the data after the join
View(joined_data)


```


### Step 3

We want to identify the top 5-6 bird species sighted at the Katherine Ordway Natural History Study Area.

```{r}
# Count sightings by species
species_counts <- table(OrdwayBirds_cleaned$SpeciesName)

# Sort species counts in descending order
sorted_species <- sort(species_counts, decreasing = TRUE)

# Display the top 10 species
top_species <- head(sorted_species, n = 10)
top_species

# Define a threshold for major species (e.g., top 5 or 6)
threshold <- 5

# Create a table with major species only
major_species <- subset(OrdwayBirds_cleaned, SpeciesName %in% names(top_species)[1:threshold])
major_species

```


### Step 4

We now want to produce a month-by-month count for each of the major species, and then construct a chart to tell the story of what time of year the various species appear.

```{r}
# Prepare glyph-ready data table called `ByMonth`
ByMonth <- major_species %>%
  group_by(Month, SpeciesName) %>%
  summarise(Count = n())

# View the `ByMonth` data table
head(ByMonth)



```

Now use the ggplot template below to generate the desired graph. (Remark you can also use esquisser or mplot to create a line chart instead of the barchart)



```{r}
# ggplot commands

ByMonth %>% 
  ggplot(aes(x = Month, y = Count, color = SpeciesName)) + 
  geom_line()

ByMonth %>% ggplot(aes(x=Month,y=Count ,fill=SpeciesName)) +
  geom_bar(stat='identity',position='stack', width=.9) 

```

According to the graph, we can answer the questions posed:

##### 1. Which species are present year-round?  
*Black-capped Chickadee is the species that is present year-round*


##### 2. Which species are migratory, that is, primarily present in one or two seasons?

*Tree-sparrow, Robin and Slate-colored Junco are primarily present in one or two seasons*


##### 3. What is the peak month for each major species? (bullet list)

```{r}
cat("Peak month for each major species:\n")
cat("* American Goldfinch-October\n")
cat("* Black-capped Chickadee-November\n")
cat("* Robin-May\n")
cat("* Slate-colored Junco-October\n")
cat("* Tree Sparro-March\n")
```


##### 4. Which major species are seen in good numbers for at least 6 months of the year?



```{r}
# Find major species seen in good numbers for at least 6 months
six_month_species <- ByMonth %>%
  group_by(SpeciesName) %>%
  filter(n_distinct(Month) >= 6) %>%
  distinct(SpeciesName)

# Print the major species seen in good numbers for at least 6 months
cat("Major species seen in good numbers for at least 6 months:\n")
cat(paste("* ", six_month_species$SpeciesName, "\n", sep = ""))

```

